Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution

### 2020-11-09 Nicolas.tex
MS and YC comments implmented

\abstract{The true nature of type Ia supernovae (SNe~Ia) remains largely
    unknown, and as survey statistics increase, the question of astrophysical
    systematic uncertainties rises, notably that of the SN~Ia population
    evolution. In this paper, we study the dependence with redshift of the SN~Ia
    \texttt{SALT2.4} lightcurve stretch, a purely intrinsic SN property, to
    probe its potential redshift drift. The SN stretch has been shown to
    strongly correlate with the SN environment, notably with stellar age
    tracers. We model the underlying stretch distribution as a function of
    redshift, using the evolution of the fraction of young and old SNe~Ia as
    predicted by \cite{rigault2018}, and assuming constant underlying stretch
    distribution for each age population made of Gaussian mixtures. We test our
    prediction against published samples \nn{\sout{chosen} cut} to have
    \yc{\sout{negligible} marginal} magnitude selection effects, so that any observed change is
    indeed of astrophysical and not observational origin. \nn{An ongoing
    analysis with the SNANA package will test this assumption in order to refine
    the study.} \yc{YC: I don't think this should be in the abstract, rather emphasize it's a preliminary study.} \yc{\sout{We \nn{\sout{clearly}} demonstrate} There are indications} that the underlying SN~Ia stretch
    distribution is evolving as a function of redshift, and that the young/old
    drifting model is a \yc{\sout{much}} better description of the data than any time-constant
    model, including the sample-based asymmetric distributions usually used to
    correct Malmquist bias. The favored underlying stretch model is the bimodal one
    derived from \cite{rigault2018}: a high-stretch mode shared by both young and
    old environments, and a low-stretch mode exclusive to old environments. The
    precise impact of the redshift evolution of the SN Ia population intrinsic
    properties on cosmology remains to be studied. Yet, the astrophysical drift of
    the SN stretch distribution does affect current Malmquist bias corrections and
    thereby distances derived from SN affected by selection effects. We highlight
    that such a bias will increase with surveys covering increasingly larger
    redshift ranges, which is particularly important for LSST.}

@@

\yc{YC: you should introduce the principle of a fiducial sample, and a more conservative one, smaller and therefore less statistically constraining, but inherently less prone to selection effects. If the redshift drift is still significant in the conservative sample, it would be even more significant in a carefully-tailored selection-free sample: that's the goal of this preliminary paper.}

@@

The sample selection is summarized in Table~\ref{tab:sample}, and the redshift
distribution of these three surveys is shown in Fig.~\ref{fig:cuts}. As
expected, the selected redshift limits roughly correspond to the peak of these
histograms. \nn{We test in Section~\ref{ssec:verify} whether we can consider
this sample to be free of \yc{(or at least marginally affected by)} selection effects. The consistency of the results for the two samples, discussed} in
Section~\ref{sec:results}, further supports that this sample selection indeed
provides a subset of SNe~Ia with insignificant selection effects when
compared to state-of-the-art Malmquist correction techniques.

@@

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2018} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2; SNe~Ia time series are published in \citealt{saunders2020}, see also
\citealt{aldering2020}). The spectroscopic follow-up was done over a redshift
range of $0.02 < z < 0.09$~\citep[as in][]{rigault2018}, while the search was
much deeper. These 114 SNfactory SNe~Ia \nn{(80\% of the 141 SNe presented in
the previously-cited paper)} are thus in the volume-limited part of the
survey (Aldering et al., in prep.), and are therefore assumed to be a random
sampling of the underlying SN population. The SNfactory sample is
particularly useful for studying SN property drift, as it enables us to have
a large \yc{complete} SN~Ia sample at $z<0.1$.

@@

We present the stretch distribution and redshift histogram of these five surveys
up to their respective $z_{\lim}$ in Fig.~\ref{fig:sample}. \nn{Proportionally few low-stretch SNe (typically $x_1 < -1$) are observed at $z>0.4$}. \yc{YC: this rather looks like $z\sim 0.2-0.3$, as in Fig.~\ref{fig:modelall}.} The evolution of the mean stretch is also shown in
Fig.~\ref{fig:modelall}, where the data are split in redshift bins of regular
sample size. We see that SNe~Ia at higher redshift have on average larger
stretch ($0.34 \pm 0.10$ at $z\sim0.65$) than those at lower redshift ($-0.17\pm
0.10$ at $z\sim0.05$), suggesting that the underlying stretch distribution is
drifting. \yc{YC: I prefer ``evolving with redshift''.}

@@

In this analysis, we have built a volume-limited sample from a set of magnitude
limited ones, using a simple redshift limit. \nn{The advantage of this approach is that our results are robust to any potential inaccuracy in the modeling of the complex
selection functions for the individual datasets. \yc{YC: once again, I don't think it's a fair statement. You should rather emphasize that this drastic approach is more robust at the price of being statistically sub-optimal, but is enough to demonstrate the key finding: a redshift evolution of stretch compatible with R+18 model.} However, while we will find consistent results when considering a sample derived from conservative cuts (Section~\ref{sec:results}), there remains the possibility that our sample is not fully volume-limited and that unaccounted for selection effects remain. To fully account for this possibility, a} follow up analysis will make use of the SNANA package \citep{SNANA} to account
for more subtle selection functions; \nn{however}, it is still necessary to test
if the sample we have created could show sign of leftover sample selection.
%While our conservative comparison sample is That fact lead to the use of conservative cuts as first indicators that there isa drift that can't be explained by selection effects alone and a}n ongoing 
\yc{YC: this paragraph to be rephrased IMHO.}

To do so, we compare the stretch and color distributions of the SNe~Ia in our
sample between datasets of overlapping redshifts. If indeed we have built a
volume-limited sample, the distributions of any SNe~Ia parameter at a given
redshift should not depend of their survey or origin. \yc{YC: survey/origin, what's the difference?} When considering a
redshift range, this has to be narrow enough such that any drift would be
negligible. \yc{YC: why do you say that here, since you didn't fine-tuned the overlaping $z$-range?} The two samples that overlap the most in redshift are PS1 and SDSS
in the redshift range $0.10 < z < 0.20$ (see Fig~\ref{fig:sample}). This
corresponds to the 146 SNe~Ia from SDSS at the highest redshift end and thus the
most affected by potential selection effect leftover (see the corresponding
discussion in section~\ref{sec:sample} \yc{YC: we are in Sec.~\ref{sec:sample}!}). In that same redshift range, PS1 has 52
SNe~Ia, which are at in the lowest redshift bin and thus unlikely to have any
selection issue. \yc{\sout{The}} Fig.~\ref{fig:distrib} present\yc{s} the stretch and color
distribution of both \yc{YC: what?}, together with $p$-values of Kolmogorov-Smirnov tests that
\yc{\sout{\nn{probe whether}}} the $x_1$ and $c$ of the SDSS and PS1 samples at $0.10 < z <
0.20$ arise from the same underlying distribution. These $p$-values are high
($>10\%$), and if one cannot conclude from this test that these SNe~Ia are from
the same underlying distribution, one has no ground to claim they are not. \yc{YC: I suggest ``These $p$-values are too high ($>10\%$) to exclude that both samples originate from the same underlying population.''} This
is in agreement with the visual impression from Fig.~\ref{fig:distrib}. \nn{We perform a}
similar analysis for PS1 and SNLS in the redshift range $0.20 < z
< 0.31$ where the same conclusion \nn{can} be drawn, i.e., there is no sign of
discrepancy in the stretch and color distributions between the low- and high-end
of the SNLS and PS1 datasets, respectively. Nonetheless, the small size of the
SNLS dataset at $z < 0.31$ (12 SNe~Ia vs. 90 for PS1) limits the interest of the
test as only strong deviation would be noticeable. Extending the redshift range
to $0.20 < z < 0.40$ (though we have no PS1 data above 0.3) enables to get 31
SNe~Ia from SNLS in total, the stretch and color \yc{KS-test} $p$-values remains \yc{\sout{high} inconclusive}, at 34\%
and 4\% respectively.

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/both-cut_SDSS_SNLS_PS1.pdf}
    \caption{$x_1$ \yc{(left)} and $c$ \yc{(right)} distribution histograms of different surveys
        overlapping in redshift. \textit{Facing up}: SDSS and PS1 within the
        redshift range $0.10 < z < 0.20$; \textit{facing down}: PS1 and SNLS
        within the redshift range $0.20 < z < 0.31$. Error bars show the Poisson
        noise \yc{YC: not visible enough}. Our stretch ``Base'' model\yc{\sout{ing}} is illustrated in orange at the mean
        redshift of the redshift ranges, $0.15$ and $0.25$, respectively.
        \yc{Kolmogorov-Smirnov test $p$-values} are indicated on the top
        (\yc{resp.} bottom) of each panel showing no sign that the SDSS and PS1 (\yc{resp.} PS1 and
        SNLS) $x_1$ and $c$ distributions are not drawn from the same underlying
        distributions.}
    \label{fig:distrib}
\end{figure}

We \yc{furthermore} highlight that the SNe~Ia color is more prone to selection effects than
stretch as illustrated in Fig.~\ref{fig:maglim}; see also e.g., Fig.~3 of
\cite{kessler2017}. Hence, since the color \yc{distribution} shows no \yc{\sout{sign} significant hint} of \yc{\sout{leftover}} selection
effect, it further supports our claim that our simple redshift-based selection
criteria is sufficient to \yc{\sout{test the \nn{existence of the} redshift drift of the
SN~Ia stretch distribution} build an unbiased SN~Ia sample at all redshifts required to test the redshift evolution of the stretch distribution}.

@@

We report in Table~\ref{tab:bbc} \nn{our determination of} the samples' $\mu_0$
and $\sigma_{\pm}$ adjusted on the nominally selection-free samples using our
fiducial cuts (see Section~\ref{sec:sample}) \nn{when implementing an asymmetric
Gaussian fit}. \yc{YC: ???} We find our results in close agreement with \cite{scolnic2016}
for SNLS and SDSS and with \cite{scolnic2018a} for PS1, who derived these model
parameters using the full BBC formalism, using numerous simulations to model the
selection effects (see details e.g., Section~3 of \citealt{kessler2017}). The
agreement between our fit of the asymmetric Gaussians on the supposedly
selection-free part of the samples and the results derived using the BBC
formalism supports our approach to get a sample with negligible selection
effects. If we were to use \cite{scolnic2016} and \cite{scolnic2018a} best fit
values of the $\mu_0, \sigma_{\pm}$ asymmetric parameters for SNLS, SDSS and
PS1, respectively, the $\Delta$AIC between our Base drifting model and the BBC
modeling would go even deeper from $-32$ to $-47$. We further discuss the
consequence of this result for cosmology in Section~\ref{sec:discussion}.

@@

In this paper, we considered a simple \yc{two-population} Gaussian mixture modeling, but additional
data free from significant Malmquist bias would enable us to refine it as
necessary. We note that samples at the low- and high-redshift ends of the Hubble
diagram would be particularly helpful for this drifting analysis; fortunately
this will soon be provided by the Zwicky Transient Facility
\citep[low-$z$,][]{bellm2019, graham2019}, and Subaru and SeeChange SNe~Ia
programs (high-$z$), respectively. \nn{To \yc{\sout{fully break the degeneracy between} improve the disentangling of} \yc{observational} selection effects and \yc{intrinsic} population evolution, an analysis using
the SNANA framework is underway. It will allow for subtle selection effects to be \yc{\sout{considered} accounted for}, and by removing the need for a \yc{naive} redshift cut, increase the redshift leverage and statistical significance of our results.}

@@

% X
% Y
% Z
\end{thebibliography}
\end{document}end{document}

>>>>

\abstract{The true nature of type Ia supernovae (SNe~Ia) remains largely
    unknown, and as survey statistics increase, the question of astrophysical
    systematic uncertainties rises, notably that of the SN~Ia population
    evolution. In this paper, we study the dependence with redshift of the SN~Ia
    \texttt{SALT2.4} lightcurve stretch, a purely intrinsic SN property, to
    probe its potential redshift drift. The SN stretch has been shown to
    strongly correlate with the SN environment, notably with stellar age
    tracers. We model the underlying stretch distribution as a function of
    redshift, using the evolution of the fraction of young and old SNe~Ia as
    predicted by \cite{rigault2018}, and assuming constant underlying stretch
    distribution for each age population made of Gaussian mixtures. We test our
    prediction against published samples cut to have marginal magnitude
    selection effects, so that any observed change is indeed of astrophysical
    and not observational origin. \nn{In this preliminary study,} there are
    indications that the underlying SN~Ia stretch distribution is evolving as a
    function of redshift, and that the young/old drifting model is a better
    description of the data than any time-constant model, including the
    sample-based asymmetric distributions usually used to correct Malmquist
    bias. The favored underlying stretch model is the bimodal one derived from
    \cite{rigault2018}: a high-stretch mode shared by both young and old
    environments, and a low-stretch mode exclusive to old environments. The
    precise impact of the redshift evolution of the SN Ia population intrinsic
    properties on cosmology remains to be studied. Yet, the astrophysical drift
    of the SN stretch distribution does affect current Malmquist bias
    corrections and thereby distances derived from SN affected by selection
effects. We highlight that such a bias will increase with surveys covering
increasingly larger redshift ranges, which is particularly important for LSST.}

@@

\nn{We will thus consider a set of cuts that will define a first fiducial
    sample, taking the farthest limits possible; but as this solution might be
    suboptimal to create a complete sample, we also consider another set of cuts
    to define a conservative sample, smaller and therefore less statistically
    constraining, but inherently less prone to selection effects. If the
    redshift drift is still significant in the conservative sample, it would be
    even more significant in a carefully-tailored selection-free sample: the
goal of this preliminary paper is to make a first implementation of this
modeling before refining it on a better sample.}

@@

The sample selection is summarized in Table~\ref{tab:sample}, and the redshift
distribution of these three surveys is shown in Fig.~\ref{fig:cuts}. As
expected, the selected redshift limits roughly correspond to the peak of these
histograms. \nn{We test in Section~\ref{ssec:verify} whether we can consider
this sample to be free of (or at least marginally affected by) selection
effects. The consistency of the results for the two samples, discussed} in
Section~\ref{sec:results}, further supports that this sample selection indeed
provides a subset of SNe~Ia with insignificant selection effects when compared
to state-of-the-art Malmquist correction techniques.

@@

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2018} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2; SNe~Ia time series are published in \citealt{saunders2020}, see also
\citealt{aldering2020}). The spectroscopic follow-up was done over a redshift
range of $0.02 < z < 0.09$~\citep[as in][]{rigault2018}, while the search was
much deeper. These 114 SNfactory SNe~Ia \nn{(80\% of the 141 SNe presented in
the previously-cited paper)} are thus in the volume-limited part of the
survey (Aldering et al., in prep.), and are therefore assumed to be a random
sampling of the underlying SN population. The SNfactory sample is
particularly useful for studying SN property drift, as it enables us to have
a large complete SN~Ia sample at $z<0.1$.

@@

We present the stretch distribution and redshift histogram of these five surveys
up to their respective $z_{\lim}$ in Fig.~\ref{fig:sample}. \nn{Proportionally
few low-stretch SNe (typically $x_1 < -1$) are observed at $z>0.2$}. The
evolution of the mean stretch is also shown in Fig.~\ref{fig:modelall}, where
the data are split in redshift bins of regular sample size. We see that SNe~Ia
at higher redshift have on average larger stretch ($0.34 \pm 0.10$ at
$z\sim0.65$) than those at lower redshift ($-0.17\pm 0.10$ at $z\sim0.05$),
suggesting that the underlying stretch distribution is evolving with redshift.

@@

In this analysis, we have built a volume-limited sample from a set of magnitude
limited ones, using a simple redshift limit. \nn{This drastic approach is robust
at the price of being statistically sub-optimal, but is enough to
demonstrate our key finding: a redshift evolution of stretch compatible with
the \cite{rigault2018} model.} However, while we will find consistent results
when considering a sample derived from conservative cuts
(Section~\ref{sec:results}), there remains the possibility that our sample is
not fully volume-limited and that unaccounted for selection effects remain. To
fully account for this possibility, a follow up analysis will make use of the
SNANA package \citep{SNANA} to account for more subtle selection functions;
\nn{yet, it appears to still be necessary} to test if the sample we have created
could show sign of leftover sample selection.

To do so, we compare the stretch and color distributions of the SNe~Ia in our
sample between datasets of overlapping redshifts. If indeed we have built a
volume-limited sample, the distributions of any SNe~Ia parameter at a given
redshift should not depend of their survey of origin. When considering a
redshift range, this has to be narrow enough such that any drift would be
negligible. \yc{YC: \textit{why do you say that here, since you didn't fine-tune
the overlaping $z$-range?}} \nn{\textit{Because we still took that into
account?}} The two samples that overlap the most in redshift are PS1 and SDSS in
the redshift range $0.10 < z < 0.20$ (see Fig~\ref{fig:sample}). This
corresponds to the 146 SNe~Ia from SDSS at the highest redshift end and thus the
most affected by potential selection effect leftover (see the corresponding
discussion in section~\ref{ssec:cuts}). In that same redshift range, PS1 has 52
SNe~Ia, which are in the lowest redshift bins and thus unlikely to have any
selection issue. Fig.~\ref{fig:distrib} presents the stretch and color
distribution of both \nn{these surveys}, together with $p$-values of
Kolmogorov-Smirnov tests that \yc{\sout{\nn{probe whether}}}\nn{\textit{I don't
see anything else to put here without repeating ``test''}} the $x_1$ and $c$ of
the SDSS and PS1 samples at $0.10 < z < 0.20$ arise from the same underlying
distribution. These $p$-values are high ($>10\%$), and if one cannot conclude
from this test that these SNe~Ia are from the same underlying distribution, one
has no ground to claim they are not. \yc{YC: I suggest ``These $p$-values are
too high ($>10\%$) to exclude that both samples originate from the same
underlying population.''} This is in agreement with the visual impression from
Fig.~\ref{fig:distrib}. We perform a similar analysis for PS1 and SNLS in the
redshift range $0.20 < z < 0.31$ where the same conclusion can be drawn, i.e.,
there is no sign of discrepancy in the stretch and color distributions between
the low- and high-end of the SNLS and PS1 datasets, respectively. Nonetheless,
the small size of the SNLS dataset at $z < 0.31$ (12 SNe~Ia vs. 90 for PS1)
limits the interest of the test as only strong deviation would be noticeable.
Extending the redshift range to $0.20 < z < 0.40$ (though we have no PS1 data
above 0.3) enables to get 31 SNe~Ia from SNLS in total, yet the stretch and
color KS-test $p$-values remains inconclusive, at 34\% and 4\% respectively.

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/both-cut_SDSS_SNLS_PS1.pdf}
    \caption{$x_1$ (left) and $c$ (right) distribution histograms of different
        surveys overlapping in redshift. \textit{Facing up}: SDSS and PS1 within
        the redshift range $0.10 < z < 0.20$; \textit{facing down}: PS1 and SNLS
        within the redshift range $0.20 < z < 0.31$. Error bars show the Poisson
        noise \yc{YC: not visible enough}. Our stretch ``Base'' model is
        illustrated in orange at the mean redshift of the redshift ranges,
        $0.15$ and $0.25$, respectively.  Kolmogorov-Smirnov test $p$-values are
    indicated on the top (resp. bottom) of each panel showing no sign that the
SDSS and PS1 (resp. PS1 and SNLS) $x_1$ and $c$ distributions are not drawn from
the same underlying distributions.}
    \label{fig:distrib}
\end{figure}

We furthermore highlight that the SNe~Ia color is more prone to selection
effects than stretch as illustrated in Fig.~\ref{fig:maglim}; see also e.g.,
Fig.~3 of \cite{kessler2017}. Hence, since the color distribution shows no
significant hint of selection effect, it further supports our claim that our
simple redshift-based selection criteria is sufficient to \nn{build an unbiased
SNe~Ia sample at all redshifts required to test the redshift evolution of the
stretch distribution}.

@@

We report in Table~\ref{tab:bbc} \nn{our determination of} the samples' $\mu_0$
and $\sigma_{\pm}$ \nn{when implementing an asymmetric Gaussian model}, adjusted
on the nominally selection-free samples using our fiducial cuts (see
Section~\ref{sec:sample}). We find our results in close agreement with
\cite{scolnic2016} for SNLS and SDSS and with \cite{scolnic2018a} for PS1, who
derived these model parameters using the full BBC formalism, using numerous
simulations to model the selection effects (see details e.g., Section~3 of
\citealt{kessler2017}). The agreement between our fit of the asymmetric
Gaussians on the supposedly selection-free part of the samples and the results
derived using the BBC formalism supports our approach to get a sample with
negligible selection effects. If we were to use \cite{scolnic2016} and
\cite{scolnic2018a} best fit values of the $\mu_0, \sigma_{\pm}$ asymmetric
parameters for SNLS, SDSS and PS1, respectively, the $\Delta$AIC between our
Base drifting model and the BBC modeling would go even deeper from $-32$ to
$-47$. We further discuss the consequence of this result for cosmology in
Section~\ref{sec:discussion}.

@@

In this paper, we considered a simple two-population Gaussian mixture modeling,
but additional data free from significant Malmquist bias would enable us to
refine it as necessary. We note that samples at the low- and high-redshift ends
of the Hubble diagram would be particularly helpful for this drifting analysis;
fortunately this will soon be provided by the Zwicky Transient Facility
\citep[low-$z$,][]{bellm2019, graham2019}, and Subaru and SeeChange SNe~Ia
programs (high-$z$), respectively. \nn{To improve the disentangling of
observational} selection effects and \nn{intrinsic} population evolution, an
analysis using the SNANA framework is underway. It will allow for subtle
selection effects to be \nn{accounted for}, and by removing the need for a
\nn{naive} redshift cut, increase the redshift leverage and statistical
significance of our results.

@@

% X
% Y
% Z
\end{thebibliography}
\end{document}
